---
title: "Climate Growth Analysis for Whitebark Pine"
author: "Ceci Martinez"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \renewcommand*\familydefault{\sfdefault} %% this picks a sans serif font
- \usepackage[T1]{fontenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(class.source = 'number-lines', echo = TRUE)
```

## Load Packages

```{r, message=FALSE, warning=FALSE, include = F}

library(here)       # for relativizing file paths 
library(Matrix)     # for handling matrices 
library(lme4)       # for running mixed effects models
library(mgcv)       # for running GAMS
library(gratia)     # for visualizing GAM models with ggplot functionality
library(readr)      # for loading in data
library(ggplot2)    # for plotting 
library(ggmap)      # for plotting maps
library(dplR)       # for tree ring data specific analaysis
library(tidyr)      # for data wrangling 
library(dplyr)      # for data wrangling 
library(performance)
library(broom)

```

## TO DO LIST

- get kichas data in the model and see how results change
- 




## Load Data

```{r, message=FALSE, warning=FALSE}

dat_all <- read_csv(here::here("tree-H", "data", "processed", "final_df.csv"))

unique(as.factor(dat_all$dataset))



```


## Model Building

### fixed effects models
```{r, message=FALSE, warning=FALSE}

# model with MAP/MAT aggregated time varying precip and previous summer tmax
mod_climate_01 <- lm(log(RW + 0.001) ~ Z + I(Z^2) + pJulSep_JulSep_ppt + pJulAug_tmax + precip + meantemp, 
     data = dat_all)
summary(mod_climate_01) 
check_model(mod_climate_01) ###Best model with AIC

mod_climate_01_sq <- lm(log(RW + 0.001) ~ Z + pJulSep_JulSep_ppt + pJulAug_tmax + precip + meantemp, 
     data = dat_all)
summary(mod_climate_01_sq) 
check_model(mod_climate_01_sq) ###Best model with AIC

# all model terms, no quadratic for size
mod_climate_02 <- lm(log(RW + 0.001) ~ Z + JulAug_tmax + pJulAug_tmax + AprMay_tmax + pAprMay_tmax +
                         JulSep_ppt + pJulSep_ppt + pJulSep_JulSep_ppt + precip + meantemp, 
                     data = dat_all)
summary(mod_climate_02)
check_model(mod_climate_02)
confint(mod_climate_02, method = "Wald")

mod_climate_02_sq <- lm(log(RW + 0.001) ~ Z + I(Z^2) + JulAug_tmax + pJulAug_tmax + AprMay_tmax + pAprMay_tmax +
                         JulSep_ppt + pJulSep_ppt + pJulSep_JulSep_ppt + precip + meantemp, 
                     data = dat_all)
summary(mod_climate_02_sq)
check_model(mod_climate_02_sq)
confint(mod_climate_02_sq, method = "Wald")

#only quadratic term 
mod_climate_03_sq <- lm(log(RW + 0.001) ~ Z + I(Z^2) + pJulSep_JulSep_ppt + pJulAug_tmax + pAprMay_tmax + precip + meantemp, 
                     data = dat_all)
summary(mod_climate_03_sq) 

#only linear term for size
mod_climate_03 <- lm(log(RW + 0.001) ~ Z + pJulSep_JulSep_ppt + pJulAug_tmax + pAprMay_tmax + precip + meantemp, 
                     data = dat_all)
summary(mod_climate_03) 
check_model(mod_climate_03) ###Best model with AIC
```
### checking, and plotting fixed effects models 
```{r}
############## Checking models with AIC ######################## 
AIC(mod_climate_01, mod_climate_01_sq, mod_climate_02, mod_climate_02_sq, mod_climate_03, mod_climate_03_sq)
 #model 2sq has lowest AIC

#best model with AIC is Model 4
model_summary_02_sq <- tidy(mod_climate_02_sq, conf.int = TRUE) %>% 
  filter(term != "(Intercept)")

ggplot(model_summary_02_sq, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "cadetblue") +
  coord_flip() +  # Flip coordinates for a horizontal plot
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1) +
  labs(x = "Coefficients", y = "Estimate (with 95% CI)") +
  theme_minimal()

```

### mixed effects models

```{r, message=FALSE, warning=FALSE}


variable_ppt <- c("JulSep_ppt", "pJulSep_ppt", "pJulSep_JulSep_ppt", "prevJun_currAug_ppt")
variable_tmax <- c("JulAug_tmax", "pJulAug_tmax", "AprMay_tmax", "pAprMay_tmax")
f4 <- as.formula(paste("log(RW + 0.001) ~ Z + I(Z^2) + meantemp + precip", paste(c(variable_ppt, variable_tmax), collapse = " + "), sep = " + "))

# terms with only linear size
mod_climate_RE <- lmer(
  log(RW + 0.001) ~ Z  + pJulAug_tmax + pAprMay_tmax +
                           pJulSep_JulSep_ppt + precip + meantemp + 
    (1 + Z | TRE_CN),
  data = dat_all,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

mod_climate_RE@beta
summary(mod_climate_RE)

# terms with quadratic and lienar size

mod_climate_R_sq <- lmer(
  log(RW + 0.001) ~ Z + I(Z^2) + pJulAug_tmax + AprMay_tmax  +
                           pJulSep_JulSep_ppt + precip + meantemp + 
    (1 + Z | TRE_CN),
  data = dat_all,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

summary(mod_climate_R_sq)


```


### check model and viz mixed effects mdoels

```{r}
library(broom.mixed)
anova(mod_climate_RE, mod_climate_R_sq) # squared Z is better model with AIC

model_summary_RE <- broom.mixed::tidy(mod_climate_RE, conf.int = TRUE) %>% 
  filter(effect == "fixed", term != "(Intercept)")

# Plot the coefficients for mod_climate_RE
ggplot(model_summary_RE, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "cadetblue") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  labs(title = "Coefficient Plot: mod_climate_RE",
       x = "Coefficient",
       y = "Estimate (with 95% CI)") +
  theme_minimal()


# Tidy the mod_climate_R_sq model with confidence intervals
model_summary_R_sq <- broom.mixed::tidy(mod_climate_R_sq, conf.int = TRUE) %>% 
  filter(effect == "fixed", term != "(Intercept)")

# Plot the coefficients for mod_climate_R_sq
ggplot(model_summary_R_sq, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "cadetblue") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  labs(title = "Coefficient Plot: mod_climate_R_sq",
       x = "Coefficient",
       y = "Estimate (with 95% CI)") +
  theme_minimal()

```
### plotting growth functions linear and quadratic on data

```{r}


z_plot <- seq(min(dat_all$Z, na.rm = TRUE), max(dat_all$Z, na.rm = TRUE), length.out = 100)
size_plot <- z_plot * mod_climate_RE@beta[2]
size_plot_sq <- z_plot * mod_climate_R_sq@beta[2] + z_plot^2 * mod_climate_R_sq@beta[3]
plot(size_plot ~ z_plot)

plot(log(dat_all$RW + 0.001) ~ dat_all$Z)
lines(size_plot~z_plot, col = "red")
lines(size_plot_sq~z_plot, col = "blue")


```
### mixed effects model with interaction terms

```{r}

data_FIA <- dat_all %>% 
  filter(dataset == "FIA")
data_kichas <- dat_all %>% 
  filter(dataset == "kichas")

mod_climate_RE_int <- lmer(
  log(RW + 0.001) ~ Z +
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = dat_all,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)


mod_climate_RE_int_sq <- lmer(
  log(RW + 0.001) ~ Z + I(Z^2) + 
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = dat_all,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

mod_climate_RE_int_FIA <- lmer(
  log(RW + 0.001) ~ Z +
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = data_FIA,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)


mod_climate_RE_int_sq_FIA <- lmer(
  log(RW + 0.001) ~ Z + I(Z^2) + 
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = data_FIA,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

mod_climate_RE_int_kichas <- lmer(
  log(RW + 0.001) ~ Z +
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = data_kichas,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)


mod_climate_RE_int_sq_kichas <- lmer(
  log(RW + 0.001) ~ Z + I(Z^2) + 
    precip * pJulSep_JulSep_ppt +
    meantemp * pJulAug_tmax +
    meantemp * pAprMay_tmax +
    (1 + Z | TRE_CN),
  data = data_kichas,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

summary(mod_climate_RE_int)

```

### viz of mixed effects model with interaction terms

```{r}

# anova(model_summary_RE_int, model_summary_RE_int_sq) # squared Z is better model with AIC

model_summary_RE_int_summary <- broom.mixed::tidy(mod_climate_RE_int, conf.int = TRUE) %>% 
  filter(effect == "fixed", term != "(Intercept)")

# Create the coefficient plot
ggplot(model_summary_RE_int_summary, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "cadetblue") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  labs(title = "Coefficient Plot: mod_climate_RE_int",
       x = "Coefficient",
       y = "Estimate (with 95% CI)") +
  theme_minimal()

model_summary_RE_int_sq_summary <- broom.mixed::tidy(mod_climate_RE_int_sq, conf.int = TRUE) %>% 
  filter(effect == "fixed", term != "(Intercept)")

# Create the coefficient plot
ggplot(model_summary_RE_int_sq_summary, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "cadetblue") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  labs(title = "Coefficient Plot: mod_climate_RE_int_sq",
       x = "Coefficient",
       y = "Estimate (with 95% CI)") +
  theme_minimal()

z_plot <- seq(min(dat_all$Z, na.rm = TRUE), max(dat_all$Z, na.rm = TRUE), length.out = 100)
size_plot <- z_plot * mod_climate_RE_int@beta[2]
size_plot_sq <- z_plot * mod_climate_RE_int_sq@beta[2] + z_plot^2 * mod_climate_RE_int_sq@beta[3]
plot(size_plot ~ z_plot)

plot(log(dat_all$RW + 0.001) ~ dat_all$Z)
lines(size_plot~z_plot, col = "red")
lines(size_plot_sq~z_plot, col = "blue")

range(log(dat_all$RW + 0.001))
dat_all <- dat_all %>% 
  filter(log(RW + 0.001) < 2)


```
Warmer year, warm location, less growth
Cooler years, cool locations, less growth
Warm year, cool location, higher than expected growth?? 

### plotting interaction terms 

```{r}
# start with single tree at single location 


dat_onetree <- dat_all %>% 
  # filter(TRE_CN == "22T1778") %>% 
  dplyr::select(-c("pJulAug_tmax", "meantemp")) # for each row give meantemp, and maxtemp


int_grid <- expand_grid(meantemp = seq(-3,3, length.out = 3), pJulAug_tmax = seq(-3,3, length.out = 3))


dat_pred <- do.call(rbind, lapply(1:nrow(int_grid), function(i) dat_onetree %>% 
         mutate(meantemp = int_grid[i,]$meantemp, pJulAug_tmax = int_grid[i,]$pJulAug_tmax)))

dat_pred <- dat_pred %>% 
  drop_na(Z)

nrow(dat_pred)
dat_pred <- dat_pred %>% 
  mutate(y_pred = predict(mod_climate_RE_int, newdata = dat_pred))

y_pred = predict(mod_climate_RE_int, newdata = dat_pred)
length(y_pred) #10,387

nrow(dat_all) #20,450
summary(dat_all)

dat_pred %>% filter(TRE_CN %in% sample(unique(TRE_CN), 10)) %>%
 ggplot() + geom_line(aes(y = y_pred, x = pJulAug_tmax, group = year)) + 
  facet_grid(meantemp~TRE_CN)


```
